---
title: "compare_QAPA_outputs"
output: html_notebook
---

Markdown setup and load libraries

```{r setup, warning=FALSE}
library(dplyr)
library(data.table)
library(stringr)
library(tidyr)
library(ggplot2)
library(qvalue)
```

Load QAPA output

```{r load_qapa_output}
qapa_output = fread("../pau_results.txt")
qapa_output = qapa_output %>% replace(is.na(.), 0)
```

Find the comparison we are currently doing

```{r get_comparison}
comparison = as.character(str_split(basename(getwd()), "_vs_")[[1]])
```

Link this to the file names (from file_naming.tsv)

```{r link_file_names}
file_namings = fread("../../../file_naming.tsv", header=F)

file_namings = file_namings %>% filter(V1 %in% comparison)
condition1_files = list.files("../../quantification", pattern=as.character(file_namings[1,2]))
condition2_files = list.files("../../quantification", pattern=as.character(file_namings[2,2]))

condition1_filenames = sapply(condition1_files, function(x) substr(x, 0, nchar(x)-5)) %>% unname()
condition2_filenames = sapply(condition2_files, function(x) substr(x, 0, nchar(x)-5)) %>% unname()
```

Filter qapa_output columns by those relevant to this comparison
N.b. we want to keep the first 12 columns so set these to true after

```{r filter_comparison_columns}
column_filter = grepl(paste0(paste0(condition1_filenames, collapse="|"),"|",paste0(condition2_filenames, collapse="|")), colnames(qapa_output))
column_filter[1:12]=TRUE
qapa_output = as.data.frame(qapa_output)
qapa_output = qapa_output[,column_filter]
```

Extract condition-condition specific columns into 2 separate DFs. Further split this into TPM and PAU DFs.

```{r split_df}
condition1_cols = qapa_output[,grepl(paste0("APA_ID", "|", paste0(condition1_filenames, collapse="|")), colnames(qapa_output))]
condition2_cols = qapa_output[,grepl(paste0("APA_ID", "|", paste0(condition2_filenames, collapse="|")), colnames(qapa_output))]

condition1_tpm = condition1_cols[,grepl(paste0("APA_ID", "|",paste0("TPM")), colnames(condition1_cols))]
condition1_pau = condition1_cols[,grepl(paste0("APA_ID", "|",paste0("PAU")), colnames(condition1_cols))]
  
condition2_tpm = condition2_cols[,grepl(paste0("APA_ID", "|",paste0("TPM")), colnames(condition2_cols))]
condition2_pau = condition2_cols[,grepl(paste0("APA_ID", "|",paste0("PAU")), colnames(condition2_cols))]
```

Summary statistics on all of these. Also save a list of the raw PAU values so we can do a t-test later.

```{r tpm_pau_summary_stats}
#get the position of the last column with quant info in it
info_cols = dim(condition1_tpm)[2]
#rowwise summary of column values
condition1_tpm_summary = condition1_tpm %>% rowwise() %>%
  transmute(APA_ID=APA_ID, 
            median=median(c_across(2:info_cols)), 
            mean=mean(c_across(2:info_cols)))
#rename the file_names to remove replicate number
condition1_name = sapply(condition1_filenames, function(x) substr(x, 0, nchar(x)-4)) %>% unname() %>% unique()
#create vector for column headers, using the group name
col_names = c("APA_ID", 
              paste0(condition1_name, "_medianTPM"), 
              paste0(condition1_name, "_meanTPM"))
colnames(condition1_tpm_summary) = col_names

### REPEAT THIS FOR THE OTHER 3 DFs
#condition1_PAU
info_cols = dim(condition1_pau)[2]
condition1_pau_summary = condition1_pau %>% rowwise() %>%
  transmute(APA_ID=APA_ID, 
            median=median(c_across(2:info_cols)), 
            mean=mean(c_across(2:info_cols)), 
            raw_list=list(c_across(2:info_cols)))
condition1_name = sapply(condition1_filenames, function(x) substr(x, 0, nchar(x)-4)) %>% unname() %>% unique()
col_names = c("APA_ID", 
              paste0(condition1_name, "_medianPAU"), 
              paste0(condition1_name, "_meanPAU"),
              paste0(condition1_name, "_rawPAUValues"))
colnames(condition1_pau_summary) = col_names

#condition2_TPM
info_cols = dim(condition2_tpm)[2]
condition2_tpm_summary = condition2_tpm %>% rowwise() %>%
  transmute(APA_ID=APA_ID, 
            median=median(c_across(2:info_cols)), 
            mean=mean(c_across(2:info_cols)))
condition2_name = sapply(condition2_filenames, function(x) substr(x, 0, nchar(x)-4)) %>% unname() %>% unique()
col_names = c("APA_ID", 
              paste0(condition2_name, "_medianTPM"), 
              paste0(condition2_name, "_meanTPM"))
colnames(condition2_tpm_summary) = col_names

#condition2_PAU
info_cols = dim(condition2_pau)[2]
condition2_pau_summary = condition2_pau %>% rowwise() %>%
  transmute(APA_ID=APA_ID, 
            median=median(c_across(2:info_cols)), 
            mean=mean(c_across(2:info_cols)), 
            raw_list=list(c_across(2:info_cols)))
condition2_name = sapply(condition2_filenames, function(x) substr(x, 0, nchar(x)-4)) %>% unname() %>% unique()
col_names = c("APA_ID", 
              paste0(condition2_name, "_medianPAU"), 
              paste0(condition2_name, "_meanPAU"),
              paste0(condition2_name, "_rawPAUValues"))
colnames(condition2_pau_summary) = col_names
```

Merge back into dataframe with the original 12 metadata columns

```{r merge_summary_df}
summary_df = qapa_output[,1:12]
summary_df = left_join(summary_df, condition1_pau_summary, by="APA_ID") %>%
  left_join(., condition2_pau_summary, by="APA_ID") %>%
  left_join(., condition1_tpm_summary, by="APA_ID") %>%
  left_join(., condition2_tpm_summary, by="APA_ID")
```

Histogram to depict the number of possible APA events per gene

```{r histogram1}
ggplot(summary_df, aes(x=Num_Events)) + geom_histogram(binwidth=1, color="darkblue", fill="lightblue") + 
  labs(title="Possible APA events based on reference",
         y="Number of genes",
         x="Number of APA events per gene")
```

Filter for those with TPM>1 (in at least 1 condition), and PAU>0 (in both conditions)

```{r filter_expressed}
filtered_summary_df = summary_df %>% filter(.[[13]]>0 & .[[16]]>0 & (.[[19]]>1 | .[[21]]>1))
```

Update the "Num_Events" column to take into account those dropped

```{r update_num_events}
count_events = filtered_summary_df %>% group_by(Gene_Name) %>% summarize(new_count=n())
filtered_summary_df = left_join(filtered_summary_df, count_events, by="Gene_Name") %>%
  mutate(Num_Events = new_count) %>%
  select(-new_count)
```


Histogram to depict the APA events of expressed transcripts (TPM>1)

```{r histogram2}
ggplot(filtered_summary_df, aes(x=Num_Events)) + geom_histogram(binwidth=1, color="darkblue", fill="lightblue") + 
  labs(title="Number of APA events",
       subtitle="Based on expressed transcripts (TPM>1)",
         y="Number of transcripts",
         x="Number of APA events per gene") + 
  stat_bin(aes(y=..count.., label=ifelse(..count..==0,"",..count..)), geom="text", vjust=-.5) 
```

In order to determine differential APA, we need to compare the raw value (from raw_list column) data and compare the 2 conditions via a t-test, and get p-values to add to the summary table. We will call these stats diff_PAU (for "differential PolyA site usage"). And we will get a p-value, q-value, and an effect size (diff between medians).

```{r calc_diff_PAU_stats, warning=FALSE, message=FALSE}
len = nrow(filtered_summary_df)
i = 1

filtered_summary_df[,23] = 0
colnames(filtered_summary_df)[23] = "diff_PAU_pval"
suppressWarnings(
while(i<=len){
  condition1_vals = filtered_summary_df[i,15] %>% unlist() %>% unname()
  condition2_vals = filtered_summary_df[i,18] %>% unlist() %>% unname()
  
  #if we try do a t.test on 2 lists which have all values as 100, sd will = 0
  #and t.test will not be able to resolve division by 0, therefore gives error
  #therefore we will manually check if all vals are 100, and set P=1 in those
  
  if(all(condition1_vals == 100)&(all(condition2_vals == 100))){
    p_val=1
  }else{
  p_val = as.numeric(t.test(condition1_vals, condition2_vals)$p.value)
  }
  filtered_summary_df[i,23] = p_val
  i=i+1
})

filtered_summary_df[,23] = as.numeric(filtered_summary_df[,23])

#we also need to find q values (adjusted p values) to take into account multiple testing
pvals = filtered_summary_df$diff_PAU_pval
qobj = qvalue(p=pvals)
qvals = qobj$qvalues

filtered_summary_df[,24] = qvals
colnames(filtered_summary_df)[24]="diff_PAU_qval"
filtered_summary_df[,24] = as.numeric(filtered_summary_df[,24])

filtered_summary_df = filtered_summary_df %>% mutate(diff_PAU_effectsize = abs(.[[13]] - .[[16]]))
```

Now filter this for significance (based on p and q)

```{r filter_significance}
significant_APA = filtered_summary_df %>% filter(diff_PAU_pval<0.05 & diff_PAU_qval<0.05)
```

We also want to filter by effect size.

But before doing that, we can visualize how many significant differential APA events have occurred, and by what magnitude

```{r threshold_graph}
bins = seq(0,95, by=5)
meet_threshold = list()

for(cutoff in bins){
  n_meets_threshold = significant_APA %>% 
    filter(diff_PAU_effectsize>cutoff) %>%
    nrow()
  meet_threshold = append(meet_threshold, n_meets_threshold)
}

df = data.frame(bins, (meet_threshold %>% unlist()))
colnames(df) = c("threshold", "met")

ggplot(df, aes(x=threshold, y=met)) + geom_col(colour="darkblue", fill="lightblue") + 
  scale_x_continuous(breaks=seq(0,95,5)) + 
  geom_text(aes(label=met), position=position_dodge(width=0.9), vjust=-0.25) + 
  labs(title="Quantification of differential APA events with effect sizes greater than X",
       subtitle="Based on expressed transcripts (TPM>1)",
         y="Number of APA events",
         x="X (Effect size threshold)")
```

Can also visualize the effect sizes as a histogram instead of based on thresholds

```{r histogram3}
ggplot(significant_APA, aes(x=diff_PAU_effectsize)) + geom_histogram(binwidth=10,color="darkblue", fill="lightblue") + 
  labs(title="Effect size distribution",
       subtitle="Based on expressed transcripts (TPM>1) with significant differential APA (P<0.05, q<0.05)",
         y="Number of APA events",
         x="Effect size (%)") + 
  stat_bin(aes(y=..count.., label=ifelse(..count..==0,"",..count..)), geom="text", vjust=-0.5, binwidth=10) 
```
Before filtering on effect size, how many genes does this correspond to?

```{r histogram4}
num_of_events_per_gene = significant_APA %>% group_by(Gene) %>% summarize(Gene_Name=Gene_Name, sig_events=n())
#also join this to the significant_APA list to replace num_events
significant_APA = left_join(significant_APA, num_of_events_per_gene, by="Gene_Name") %>% relocate(sig_events, .after=Num_Events) %>% select(-Num_Events, -Gene.y) %>% distinct()
colnames(significant_APA)[3]="Gene"

ggplot(significant_APA, aes(x=sig_events)) + geom_histogram(binwidth=1,color="darkblue", fill="lightblue") + 
  labs(title="Number of significant differential APA events per gene",
       subtitle="Based on expressed transcripts (TPM>1) with significant differential APA (P<0.05, q<0.05)",
         y="Number of genes",
         x="Number of significant differential APA events") + 
  stat_bin(aes(y=..count.., label=ifelse(..count..==0,"",..count..)), geom="text", vjust=-0.25, binwidth=1) 
```
Other statistics:

```{r other_summary_stats}
other_stats = significant_APA %>% summarize(average_abs_effect_size = median(diff_PAU_effectsize), 
                                            average_abs_TPM_change = median(abs(.[[19]]-.[[21]])))
other_stats
```

```{r export_raw_sig_DAPA}
file_name = paste0(comparison[1], "_vs_", comparison[2], "_sig_dif_APA.csv")
export_sig_APA = significant_APA[,c(-15,-18)]
write.csv(export_sig_APA, file = file_name)
```

Filter by effect size (based on the pipeline.yml)

```{r read_config}
pipeline_yml = yaml::read_yaml("../../../pipeline.yml")
effectsize_threshold = pipeline_yml$diff_apa$minimum_effect_size

filtered_significant_APA = significant_APA %>% filter(diff_PAU_effectsize > effectsize_threshold)
```

New histogram and update num_events

```{r histogram5}
#we will call those which meet both statistical significance, and effectsize threshold 'sig2'
#and those which only meet statistical significance will be 'sig'
num_of_events_per_gene = filtered_significant_APA %>% group_by(Gene) %>% summarize(Gene_Name=Gene_Name, sig2_events=n())
#also join this to the significant_APA list to replace num_events
filtered_significant_APA = left_join(filtered_significant_APA, num_of_events_per_gene, by="Gene_Name") %>% relocate(sig2_events, .after=sig_events) %>% select(-Gene.y) %>% distinct()
colnames(filtered_significant_APA)[3]="Gene"

ggplot(num_of_events_per_gene, aes(x=sig2_events)) + geom_histogram(binwidth=1,color="darkblue", fill="lightblue") + 
  labs(title="Number of significant differential APA events per gene",
       subtitle=paste0("Based on expressed transcripts (TPM>1) with significant differential APA (P<0.05, q<0.05), \n and significant effect size (>",effectsize_threshold,"%)"),
         y="Number of genes",
         x="Number of significant differential APA events") + 
  stat_bin(aes(y=..count.., label=ifelse(..count..==0,"",..count..)), geom="text", vjust=2, binwidth=1) 
```